Joint estimation of SNP effects improves the prediction of complex diseases // Predicting complex diseases: performance and robustness [OF NEW METHOD?]

### 2018-04-19 paper/draft.tex
modifs

\abstract{
Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T). It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast. However, PRS rely on non-optimal heuristics to account for correlation and reduce noise in the predictor.

@@

In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests. The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster than the C+T method. 
Moreover, predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, whose optimal value depends on disease architecture. 
The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs. 
When analyzing real SNP data from a case-control study,  penalized logistic regression provides an AUC of 89\% while the standard C+T method provides a smaller AUC of 82\%.

Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores.\\

@@

Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. 
This technique uses GWAS summary statistics only, which makes it very convenient to compute and also very fast.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. 
Yet, the predictions resulting from LMM, known e.g.\ as gBLUP, are not designed for binary traits such as a disease status and are not optimal in this setting \cite[]{abraham2013performance}. 
Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects. Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.

We recently developed two R packages, bigstatsr and bigsnpr, for efficient analysis of large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes an efficient algorithm with new implementation tricks for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of the C+T method, our implementation of penalized logistic regression and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}. 
Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} [<- IN METHODS? MAIS BIEN AUSSI DE DIRE PQ TOT..].
For the C+T model, we compare different thresholds of inclusion of SNPs.
For the logistic regression, we include two novel approaches. 
First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model. 
We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.

@@

First, we quickly discard T-Trees because they show lower predictive performance as compared to other methods, while being computationally demanding.
Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.

@@

We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}.
Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
As for simulations presented here, in order to remove the genetic structure induced by the celiac disease status, we first restrict our simulations to controls. 
Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals. 
This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').

@@

In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set (from 1000 to 6000), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace). We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$ 
and $S_\text{causal}^{(3)}$ ($S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$). 
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ 
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$. 
Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still have a marginal effect. 
For both models, the values of $y_i$ are then standardized so that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. 
It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). 
However, comparison results for T-Trees relies on 5 simulations only (and an heritability of 80\%) because this method can require several hours of computation for a single simulation. 
Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only. Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power}. 
For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods. A total of 100 simulations are run using 300 causal SNPs anywhere on the genome.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), referred to as summary statistics.
For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. 
Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{j \in S_\text{clumping}} \mathds{1}\{p_j < p_T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

@@

Finally, for the penalized logistic regression, we solve: $$\argmin_{\beta_0, \beta}(\lambda, \alpha)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse and therefore more interpretable models (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use the elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which minimizes the log loss on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected. Finally, the $K$ resulting probabilities for each individual are averaged. 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.
Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression with the CMSA procedure, as described previously. We call this method ``logit-triple'' in the results.

@@

We use Monte Carlo cross-validation to compute AUC, partial AUC, number of predictors used and execution time for the original Celiac dataset with real phenotypes.
We randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.

@@

When there are 30 causal SNPs and 80\% of heritability, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}). 
Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs. Yet, these are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).

@@

\subsection{Other methods and ``fancy'' model}

@@

In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores. 
We show that jointly estimating SNP effects using efficient penalized logistic regressions enables to achieve higher predictive performance, as reported in a previous study \cite[]{abraham2013performance}. Indeed, when disease architecture and sample size allow for high predictive performance (AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic scores in clinical settings.
The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
However, in order to accurately predict disease status for clinical interventions, best possible predictive methods should be used. The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time.

As for the C+T method itself, we recall that we report the maximum AUC of 102 thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the method \cite[]{ware2017heterogeneity}. 
On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance.

@@

Finally, in this comparative study, we did not consider the problem of population structure \cite[]{martin2017human} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}. In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population. We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}. In another study, we wish also to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.

@@

>>>>

\abstract{
Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T). It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast. However, PRS rely on non-optimal heuristics to account for correlation and to reduce noise in the predictor.

@@

In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests. The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster than the C+T method. 
Moreover, predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, whose optimal value depends on disease architecture. 
The improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs, or when sample size is increasing. 
When analyzing real SNP data from a case-control study,  penalized logistic regression provides an AUC of 89\% while the standard C+T method provides a smaller AUC of 82\%.

Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigstatsr to obtain more accurate polygenic risk scores when individual genotype data are available.\\

@@

Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. 
This technique uses GWAS summary statistics only, which makes it very convenient to compute and also very fast.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. 
Yet, the predictions resulting from LMM, known e.g.\ as ``gBLUP'', are not designed for binary traits such as a disease status and are not optimal in this setting \cite[]{abraham2013performance}. 
Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects. Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes an efficient algorithm with new implementation tricks for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of the C+T method, our implementation of penalized logistic regression and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}. 
Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} [<- IN METHODS? MAIS BIEN AUSSI DE DIRE PQ TOT..].
For the C+T model, we compare different thresholds of inclusion of SNPs.
For the logistic regression, we include two novel approaches. 
First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model. 
We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.

@@

First, we quickly discard T-Trees because this method shows lower predictive performance as compared to other methods, while being computationally demanding.
Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}.
Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
As for simulations presented here, in order to remove the genetic structure induced by the celiac disease status, we first restrict our simulations to controls. 
Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals. 
This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing'').

@@

In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or chromosome 6 only), the size of the training set (from 1000 to 6000), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace). We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$ 
and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$. 
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ 
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$. 
Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still have a marginal effect. 
For both models, the values of $y_i$ are then standardized so that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. 
It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). 
However, comparison results for T-Trees relies on 5 simulations only (and an heritability of 80\%) because this method can require several hours of computation for a single simulation. 
Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only. Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power,marquez2017multiethnic,vilhjalmsson2015modeling}. 
For this scenario, we use only the additive (``simple'') model, but continue to compare all previous different values of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods. A total of 100 simulations per combination of parameters are run using 300 causal SNPs anywhere on the genome.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS), referred to as summary statistics.
In the GWAS, a coefficient of regression (estimated effect size) is learned independently for each SNP along with a corresponding p-value. 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. 
Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

@@

Finally, for the penalized logistic regression, we solve: $$\arg\!\min_{\beta_0,~\beta}(\lambda, \alpha)\left\{  \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization}  \right\}$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use the elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure (Figure \ref{fig:CMSA}). We call this method ``logit-triple'' in the results.

@@

We use Monte Carlo cross-validation to compute AUC, partial AUC, number of predictors used and execution time for the original Celiac dataset with real phenotypes: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.

@@

When there are 30 causal SNPs and 80\% of heritability, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}). 
Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted in simulations with 3000 causal SNPs, which are cases where all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).

@@

\subsection{The other methods and the ``fancy'' model}

@@

In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores when individual genotype data are available. 
We show that jointly estimating SNP effects using efficient penalized logistic regressions enables to achieve higher predictive performance, as reported in a previous study \cite[]{abraham2013performance}. Indeed, when disease architecture and sample size allow for high predictive performance (say AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic risk scores in clinical settings.
The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not yet the case for our penalized logistic regressions. 
However, in order to accurately predict disease status for clinical interventions, best possible predictive methods should be used. The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time.

As for the C+T method itself, we recall that we report the maximum AUC of 102 thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the method \cite[]{ware2017heterogeneity}. 
On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance [TODO: COMPARAISON AVEC LE MAX DE BIGLASSO]. Usually, cross-validation is used to assess the best value for the hyper-parameter and then the model is train again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}. Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we prefer to use CMSA when there is a model to choose from a regularization path, with the possibility of adding an early stopping criterion to fasten the whole procedure.

@@

Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}. In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population. We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}. In another study, we wish also to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{simple-CMSA}}
\caption{Illustration of one turn of the Cross-Model Selection and Averaging (CMSA) procedure. First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
Secondly, in turn, each fold is considered as an inner validation set (red) and the other ($K - 1$) folds form an inner training set (blue). A ``regularization path'' of models is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed. The model that minimizes the loss on the inner validation set is selected. Finally, the $K$ resulting models are averaged. 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.}
\label{fig:CMSA}
\end{figure}
